\documentclass[10pt,final,conference,comsoc]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc} % optional
\usepackage{amsmath,amssymb,amsfonts}
\usepackage[cmintegrals]{newtxmath}
\usepackage{cite}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{bm}
\usepackage{siunitx}
\usepackage{comment}
\usepackage{mathtools}
\usepackage{url}

\newcommand{\LRT}[2]{\mathrel{\mathop\gtrless\limits^{#1}_{#2}}}
\newcommand{\cn}{\mathcal{CN}}
\newcommand{\n}{\mathcal{N}}
\newcommand{\D}{\mathcal{D}}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\graphicspath{{../main_matlab_figures/}{../matlab/}}

\DeclareUnicodeCharacter{0303}{}

\DeclareSIUnit{\sample}{S}
\DeclareSIUnit{\bits}{b}

\begin{document}

%\title{Low Complexity Methods for Joint Detection and Synchronization of TDMA Bursts}
\title{Practical Methods for Joint Time and Carrier Synchronization in
  LPI/LPD Communications}
\author{\IEEEauthorblockN{Haotian Zhai and Bernd-Peter Paris}
\IEEEauthorblockA{\textit{Department of Electrical and Computer Engineering} \\
\textit{George Mason University}\\
Fairfax, VA 22030 \\
\{hzhai,pparis\}@gmu.edu}}

\maketitle

\begin{abstract}
  To operate at low SNR, LPI/LPD communication systems rely on
  coherent processing which requires that the receiver is precisely
  synchronized in time, frequency and phase. 
  To that end, this paper proposes a family of data-aided joint
  frame and carrier synchronization algorithms.
  A sequential detection algorithm based on the generalized likelihood ratio test (GLRT)
  is used to detect an embedded preamble signal in the received data
  stream.
  Novel algorithms for low-complexity, coarse carrier synchronization
  at low SNR provide carrier estimates
  during sequential detection.
  After detection,
  the coarse carrier estimate is refined for use in coherent demodulation.
  The proposed family of algorithms can be scaled to support operation
  over a wide range of SNR, including SNR below \SI{0}{\decibel}.
  Algorithms are validated through simulation.
  The practicality of our approach is demonstrated by real-time operation on
  a standard SDR platform with sample rates approaching
  \SI{10}{\mega\hertz}.
\end{abstract}

\begin{IEEEkeywords}
LPI/LPD, carrier synchronization, frame synchronization, sequential
detection, GLRT, SDR
\end{IEEEkeywords}

\input{introduction}

\input{model}

\input{detection}

\input{freq_est}

\input{simulations}

\input{implementation_on_SDR}

\section{Conclusion}

A family of algorithms for joint detection and carrier synchronization
suitable for LPI/LPD communications was presented.
These algorithms are designed for low complexity while achieving good
performance over a wide range of SNR.\@
We implemented our algorithms on a SDR platform and demonstrated that
they can sustain several \si{\mega\hertz} sample rate.

\bibliographystyle{ieeetr}
\bibliography{ref.bib}

\end{document}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
